Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

### 2022-01-17 analysis/paper/London_HauloutBeahvior.Rmd

  - \usepackage {sectsty}
  - \allsectionsfont {\sffamily}

@@

  - \linespread{1.1}

@@

```{r setup, include = FALSE}
library(targets)
library(conflicted)
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
library(lubridate)
conflict_prefer("month","lubridate")
library(glue)
library(scales)
library(DBI)
library(dbplyr)
library(RPostgres)
library(pingr)
library(sf)
library(flextable)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(ggtext)
library(glmmLDTS)
library(mgcv)
library(ragg)
library(stickylabeller)
library(MetBrewer)

@@

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  include = FALSE,
  fig.path = "../figures/",
  dpi = 400,
  fig.width = 6.5,
  out.width = "100%"
)

@@

Bio-logging records of time spent out of the water provide valuable data for
identifying covariates that explain variation in haul-out behavior. For
instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
bio-logging data to demonstrate greater haul-out propensity in juvenile
crabeater seals (*Lobodon carcinophaga*) than adults, with highest probabilities
in February and at times close to solar noon. In the Arctic, Von Duyke et al.
[-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly et al. [-@kelly2010] using VHF radio tags and
direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
[-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using much
larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion of
time bearded seals spent hauled out progressively increased through spring and
summer. And, Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
increased gradually starting in March and peaking in May and June for bearded,
ribbon, and spotted seals.

@@

deploy_table %>% 
  dplyr::group_by(species, sex, age) %>% 
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>% 
  dplyr::mutate(
    deploy_stats = glue::glue('{n} ({format(seal_hours, big.mark=",")} seal hours)')
  ) %>% 
  dplyr::select(-c("n","seal_hours")) %>% 
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>% 
  dplyr::ungroup() %>% 
  dplyr::rename(Species = species, Sex = sex, `Young-of-the-Year` = `YOUNG OF YEAR`,
                `Subadult` = SUBADULT, Adult = ADULT) %>% 
  flextable() %>% 
  add_header(`Young-of-the-Year` = "Age Class",
   Subadult = "Age Class", Adult = "Age Class",
   top = TRUE ) %>% 
  merge_h(part = "header") %>% 
  align(align = "left", part = "header") %>% 
  # font(fontname = "Source Serif Pro Light", part = "all") %>% 
  fontsize(size = 10, part = "all") %>% 
  autofit()
```

@@

covariate_tbl %>% 
  flextable() %>% 
  font(fontname = "Source Serif Pro Light", part = "all") %>% 
  fontsize(size = 10) %>% 
  hline(i=8) %>% 
  set_table_properties(width = .95, layout = "autofit")
```

@@

```{r beardedHOwx,fig.height = 5.5, include=TRUE, fig.id = "beardedHOwx", fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications. Hour of the day was fixed at local solar noon and day-of-year held at the day of peak haul-out for bearded seals. Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability."}

@@

```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOCal", fig.cap="Predicted haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model. Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males. 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes."}

@@

```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.id = "ribbonHOwx", fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of ribbon seals within each age and sex classification. Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for ribbon seals. Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability."}

@@

```{r spottedHOCal, fig.height = 5.5, include=TRUE, fig.id = "spottedHOCal", fig.cap="Predicted haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model. Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males. 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes."}

@@

```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.id = "spottedHOwx", fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of spotted seals within each age and sex classification. Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for spotted seals. Transparent vertical lines represent the 95% confidence interval around the predicted haul-out probability."}

@@

```{r annualHO, fig.height = 5.5, include=TRUE, fig.id = "annualHO", fig.cap="Annual variability in the timing of peak haul-out probability for ribbon and spotted seals across 10 years. Predictions are shown for local solar noon and under smoothed weather conditions. Only those groups (age-sex + year) that included observations from more than one seal are shown. Additionally, any groups where the underlying data started after 1 June or ended before 1 May are not included."}

@@

```{r maxextentHO, fig.height = 5.5, include=FALSE, fig.id = "maxextent", fig.cap="Annual timing of peak haul-out behavior analyzed against the yearly maximum area of sea ice extent in the study area. While spotted seal adult females and subadults show a negative trend line, none of the relationships were determined to be significant."}

@@

**Josh M. London**: investigation, conceptualization, methodology,
formal analysis, validation, software, writing: original draft, writing:
review and editing, visualization, and data curation

**Paul B. Conn**: conceptualization, methodology, formal analysis,
software, validation, writing: original draft, writing: review and
editing

**Stacie K. Hardy**: investigation, data curation, methodology,
validation, writing: review and editing

**Erin L. Richmond**: data curation, investigation, methodology,
validation, writing: review and editing

**Jay M. Ver Hoef**: conceptualization, methodology, software, writing:
review and editing

**Michael F. Cameron**: investigation, project administration, writing:
review and editing

**Justin Crawford**: investigation, methodology, validation, data curation, 
writing: review and editing

**Lori T. Quakenbush**: investigation, methodology, supervision, project
administration, writing: review and editing

**Andrew L. Von Duyke**: investigation, methodology, validation, data curation,
writing: review and editing

**Peter L. Boveng**: investigation, conceptualization, supervision,
project administration, writing: review and editing

>>>>

@@

  - \usepackage{setspace}
  - \onehalfspacing

@@

```{r setup, include = FALSE}
library(targets)
library(conflicted)
library(tidyverse)
conflict_prefer("filter", "dplyr")
conflict_prefer("select", "dplyr")
library(lubridate)
conflict_prefer("month","lubridate")
library(glue)
library(scales)
library(DBI)
library(dbplyr)
library(RPostgres)
library(pingr)
library(sf)
library(flextable)
library(gt)
library(patchwork)
library(rnaturalearth)
library(rnaturalearthdata)
library(ggspatial)
library(ggsci)
library(ggrepel)
library(gghighlight)
library(ggtext)
library(glmmLDTS)
library(mgcv)
library(ragg)
library(stickylabeller)
library(MetBrewer)

@@

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  include = FALSE,
  fig.path = "../figures/",
  dpi = 300,
  out.width = "100%"
)

@@

Bio-logging records of time spent out of the water provide valuable data for
identifying covariates that explain variation in haul-out behavior. For
instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
bio-logging data to demonstrate greater haul-out propensity in juvenile
crabeater seals (_Lobodon carcinophaga_) than adults, with highest probabilities
in February and at times close to solar noon. In the Arctic, Von Duyke et al.
[-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly et al. [-@kelly2010] using VHF radio tags and
direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
[-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using much
larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion of
time bearded seals spent hauled out progressively increased through spring and
summer. And, Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
increased gradually starting in March and peaking in May and June for bearded,
ribbon, and spotted seals.

@@

deploy_table %>%
  dplyr::group_by(species, sex, age) %>%
  dplyr::summarise(
    n = n(),
    seal_hours = sum(n_hours)
  ) %>%
  dplyr::mutate(
    deploy_stats = glue::glue('{n} ({format(seal_hours, big.mark=",")} seal hours)')
  ) %>%
  dplyr::select(-c("n","seal_hours")) %>%
  tidyr::pivot_wider(names_from = age, values_from = c(deploy_stats)) %>%
  dplyr::ungroup() %>%
  dplyr::rename(Species = species, Sex = sex, `Young-of-the-Year` = `YOUNG OF YEAR`,
                `Subadult` = SUBADULT, Adult = ADULT) %>%
  flextable() %>%
  add_header(`Young-of-the-Year` = "Age Class",
   Subadult = "Age Class", Adult = "Age Class",
   top = TRUE ) %>%
  merge_h(part = "header") %>%
  align(align = "left", part = "header") %>% 
  align(i = 1, align = "center", part = "header") %>% 
  fontsize(size = 10, part = "all") %>% 
  width(width = 1.35) %>% 
  width(j = ~ Species, width = 0.9) %>% 
  width(j = ~ Sex, width = 0.4)

@@

covariate_tbl %>% 
  flextable() %>% 
  fontsize(size = 10, part = "all") %>% 
  hline(i=8) %>% 
  width(width = 1.0) %>% 
  width(j = ~ Description, width = 2.5)
```

@@

```{r beardedHOwx, fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of bearded seals combined across all age and sex classifications. Hour of the day was fixed at local solar noon and day-of-year held at the day of peak haul-out for bearded seals. Transparent vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}

@@

```{r ribbonHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted haul-out probability of ribbon seals from 1 March to 15 July for each age and sex class used in the model. Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males. 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes."}

@@

```{r ribbonHOwx, fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of ribbon seals within each age and sex classification. Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for ribbon seals. Transparent vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}

@@

```{r spottedHOCal, fig.height = 5.5, include=TRUE, fig.cap="Predicted haul-out probability of spotted seals from 1 March to 15 July for each age and sex class used in the model. Adult females are indicated by 'ADULT.F' and 'ADULT.M' indicates adult males. 'SUBADULT' and 'YOUNG OF YEAR' age classes include both sexes."}

@@

```{r spottedHOwx,fig.height = 5.5, include=TRUE, fig.cap="Marginal effects of temperature, wind, pressure, and accumulated precipitation on the predicted haul-out probability of spotted seals within each age and sex classification. Hour of the day was fixed at local solar noon and day-of-year held at the average day of peak haul-out for spotted seals. Transparent vertical lines represent the 95\\% confidence interval around the predicted haul-out probability."}

@@

```{r annualHO, fig.height = 5.5, include=TRUE, fig.cap="Annual variability in the timing of peak haul-out probability for ribbon and spotted seals across 10 years. Predictions are shown for local solar noon and under smoothed weather conditions. Only those groups (age-sex + year) that included observations from more than one seal are shown. Additionally, any groups where the underlying data started after 1 June or ended before 1 May are not included."}

@@

```{r maxextentHO, fig.height = 5.5, include=FALSE, fig.cap="Annual timing of peak haul-out behavior analyzed against the yearly maximum area of sea ice extent in the study area. While spotted seal adult females and subadults show a negative trend line, none of the relationships were determined to be significant."}

@@

* **Josh M. London**: investigation, conceptualization, methodology,
formal analysis, validation, software, writing: original draft, writing:
review and editing, visualization, and data curation
* **Paul B. Conn**: conceptualization, methodology, formal analysis,
software, validation, writing: original draft, writing: review and
editing
* **Stacie K. Hardy**: investigation, data curation, methodology,
validation, writing: review and editing
* **Erin L. Richmond**: data curation, investigation, methodology,
validation, writing: review and editing
* **Jay M. Ver Hoef**: conceptualization, methodology, software, writing:
review and editing
* **Michael F. Cameron**: investigation, project administration, writing:
review and editing
* **Justin Crawford**: investigation, methodology, validation, data curation, 
writing: review and editing
* **Lori T. Quakenbush**: investigation, methodology, supervision, project
administration, writing: review and editing
* **Andrew L. Von Duyke**: investigation, methodology, validation, data curation,
writing: review and editing
* **Peter L. Boveng**: investigation, conceptualization, supervision,
project administration, writing: review and editing
